Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-26 article.tex
modif section 4 (draft, toujours en cours)

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper, at the price of parameter
tuning.

@@

@@

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions and, that it would therefore be preferrable to lower their
importance by changing appropriatly their magnitude. Additionnally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance around $t=4\text{s}$ in the figure.
Note that the particular case where there is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i) = 1$.

@@

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
partial correlation coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by \cite{}:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$.%%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, ie. $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measures between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks \cite{xxx}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performances, with respect to the metric used in the Connectomics
challenge.

Practically, the computation of coefficient $p_{i,j}$ using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components of $\Sigma$~\citep{bishop2006pattern} (with
$M$ set to $0.8 p$ in our experiments).

@@

% Datasets
We have carried out experiments with our method on the
\textit{normal}-1,2,3, and 4 datasets provided by the organizers of
the Connectomics challenge. Each of these datasets contains $p=1000$
time series of length $T=179500$ of image calcium from simulated
neuronal networks \citep{stetter2012model}. The network to infer
contains approximately 15000 edges, ie an edge density of about
1.5\%. All inference methods compared here provides a ranking of
neuron pairs according to some association score. To assess the
quality of this ranking, we use both ROC and precision-recall curves,
and we summarize them using the area under these curves, denoted
respectively AUROC and AUPRC. Only the AUROC was used in the context of the connectomics to rThe AUROC was
used to rank submissions to the challenge but the precision-recall
curve has some\citep{schrynemackers2013protocols}

Since neurons are
not self-connected ($(i, i) \not \in E, \forall i \in V$) and some
inference methods could give maximal scores to those edges,
e.g. Pearson correlation, those elements are set to have minimal
connectivity score prior assessment.

\paragraph{Effect of pre-processing}

The top of Table \ref{} reports AUPR and AUROC for all four networks
progressively introducing all filtering steps and for the last step,
the PCA based matrix inversion. All steps improves performances, with
the bigger improvement brought by the introduction of the
... Interestingly, while PCA does not really have an impact of AUROC,
it significantly increases the AUPR score.

@@

Then, we compare our method with several other methods:
\begin{itemize}
\item Standard correlation coefficient...In this case, the
  prefiltering the same as for partial correlation.
\item \textbf{GTE} \citep{stetter2012model}, the baseline of the
  challenge, which is based on transfer entrey measures for all pairs
  of neurons $i$ and $j$ the reduction of uncertainty in predicting
  future values of $i$ given the observation of the time series of the
  neuron $j$ conditionally to the observation of past time series of
  neuron $i$, whenever the average fluorescence is over a conditioning
  level. Pre-processing and conditioning level were set to reproduce
  baseline of the challenge.
\item \textbf{GENIE3}\citep{huynhthu2010inferring}: This method is a
  state-of-the-art gene network inference method \cite{dream5}. It
  translates the the inference of the network of $p$ neurons into $p$
  supervised learning tasks. For each task, the fluorescence of a
  neuron is predicted from the fluorescence of all other neurons using
  a tree based ensemble such as random forest
  \citep{breiman2001random}. The tree ensemble associates a variable
  importance score \citep{louppe2013understanding} to each neuron used
  in the predictions, which is used as a degree of associate. To reduce the computational cost of the method, we nevertheless restrained all trees to a maximum tree depth of 3.  This is likely to be detrimental The
  following parameters are used $10000$ trees, $\log_2{p}$ variables
  are randomly selected to split tree nodes and the maximal tree depth
  is constrained to 1, 2 or 3. Note that unlike the other methods, this method provides asymetric scores, meaning it can potentially predict edge directions.
\end{itemize}

Bottom of table \ref{} shows the score of these methods. ROC and PR
curves on normal-2 are shown in Figure ... Pearson correlation is the worst method

\paragraph{ROC versus PR curves.} While only ROC was , we argue here that PR scores are more appropriate in this context.

@@

The proposed approach is compared to Pearson correlation statistics
(as described in Section~\ref{sec:inference}), generalized transfer
entropy (GTE) and the GENIE3 algorithm.
\begin{itemize}
\item \textbf{GTE} \citep{stetter2012model}, the baseline of the challenge,
measures for all pairs of neurons $i$ and $j$ the reduction of uncertainty in
predicting future values of $i$ given the observation of the time series of the
neuron $j$ conditionally to the observation of past time series of neuron $i$,
whenever the average fluorescence is over a conditioning level. Pre-processing
and conditioning level were set to reproduce baseline of the challenge.

\item \textbf{GENIE3} \citep{huynhthu2010inferring} translates the inference
of the network of $p$ neurons into $p$ supervised learning tasks. For each
task, the fluorescence of a neuron is predicted from the fluorescence of all
other neurons using a tree based ensemble such as random forest
\citep{breiman2001random}. The tree ensemble associates a variable importance
score \citep{louppe2013understanding} to each neuron used in the predictions,
which is used as a degree of associate. The following parameters are used
$10000$ trees, $\log_2{p}$ variables are randomly selected to split tree nodes
and the maximal tree depth is constrained to 1, 2 or 3. \textcolor{red}{AJ Which
depth is represented on the graph?}
\end{itemize}

\subsection*{Partial correlation improves over state-of-the-art methods}

Figure \ref{fig:curves} shows the ROC curve  and precision-recall curve
for correlation, GTE and partial correlation on the \textit{normal-2}
dataset. The partial correlation outperforms other state-of-the-art methods
by at least factor of 2 in AUPRC reaching $0.347$ and an increased of $0.03$ in
AUROC. Averaging Partial correlation matrix over a wide of parameters
further improves AUPRC to $0.389$ and improve AUROC by $0.009$.
Two reasons might explain such differences: (i) only the partial
correlation-based methods and GENI3 at lesser extent distinguish
direct links from indirect one, (ii) simulated networks from the
\citep{stetter2012model} generator or neural networks in general could be
reduced to a linear model after pre-processing. Those results
generalizes over the other datasets see Appendix \textcolor{red}{REF}.

\begin{figure}[bth]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{The weighted average of partial correlation matrix, winning approach
         of the challenge, yields better than AUROC and AUPRC than all
         other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         averaged partial correlation method, partial correlation used
         filter $f_1$). Area under the curves are given between $()$
         for the corresponding methods.}
\label{fig:curves}
\end{figure}

\textcolor{red}{Only one GENI3, thus it can't be observed.}
As observed on Figure \ref{fig:curves}, the GENIE3 performance increases with
deeper trees while being still computationally feasible. The reason is twofold.
First, the deeper the tree, the larger is the conditioning. \textcolor{red}{AJ:
conditioning?} As it was observed in the comparison between correlation and
partial correlation, it screens indirect links in the inferred network.
\textcolor{red}{AJ: I don't understand the link with GENIE3} Second, the number
of small trees need to be much higher than deep trees in order to observe - and
evaluate - all pairs.  However, as suggested by \cite{louppe2014understanding},
limiting the maximal depth of trees may help to improve the variable importance
computing, i.e. inferring the degree of association. \textcolor{red}{AJ: there
is no explanation of why it would be better apart for computational reason.}
Therefore a trade-off  of the maximal depth may give the best of both worlds.
\textcolor{red}{AJ:  Which tradeoff? Which worlds?}

\subsection*{Area under the ROC curve a poor metric for network inference}

The best method only infers the \textbf{undirected} network by giving the same
degree of association to both edges $(i,j)$ and $(j,i)$. This is quite
surprising considering the problem statement. Trying to direct links could be
more penalizing than rewarding. For example, symmetrizing the degree of
association matrix does not necessarily degrade AUPRC and AUROC.

If the inferred network corresponds to the undirected network, it would
already yield high AUROC score for instance $\approx 0.996$ on
\textit{normal-1}. Infering edge directivity is likely to be counter-
productive or yield small gain. By contrast, the same experiment with AUPRC
would yield a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.

% Interestingly, methods that takes into account directivity, namely GTE and
% GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
% (in particular see GTE and GTE(undirected) in Figure \ref{fig:pr_curve}
% \textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
% especially pr-curve. AS:And now?} and Table \ref{tab:tab1} in Appendix \textcolor{red}{REF}).
% Trying to detect the link direction is more penalizing than rewarding.

The area under the ROC curve is not a consistent metric with respect to the
problem statement. A clear hierarchy is also difficult to derive from ROC
curves as shown on Figure \ref{fig:roc_curve}. We advice to avoid the AUROC for
the assessment of network inference method. An alternative would be to assess
network inference using the area under the precision-recall curve (AUPRC).
Those conclusions are in line with \cite{schrynemackers2013protocols}.

@@

In this paper, we present the winning method of the Connectomics challenge. Our
approach is simple and theoretically sounded and made of two main stages: the
signal is first preprocessed with several filters and then we derive the degree
of association matrix by computing the partial correlation statistics matrix.
We also evaluate the method in a more detailed way and perform a comparison
with other inference methods. We finally discuss the relevance of the metric
used in the challenge and we point out that it was not necessary to discover
causal relationships to maximize the ROC AUC. Thus, it may be  promising to
focus on that particular topic in future works.

@@

% We allow the bibliography to be added on top of the 6 pages.
% Supplementary material may be added in the form or a URL pointed to from
% the paper.
\bibliography{references}

@@

In our experiments, we observed that the inference process is quite sensitive about the choice of the parameter value $\tau$ and the low-pass filter. Therefore, in order to reduce the variance of the results and thus the robustness of our approach, we perform an average over a range of parameter value $\tau$ from $0.100$ to $0.209$. We also perform a weighted average over the four low-pass filters with weights 
$weight_{f_1} = 1$, $weight_{f_2} = 0.9$, $weight_{f_3} = 0.7$ and $weight_{f_4} = 0.01$, tuned to maximize the ROC AUC on \textit{normal-1}.
The weights were adjusted to take into account that some filters outperform other but there is information hold in every filter.

>>>>

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper at a price of significant
tuning.

@@

% \begin{figure}
% \centering
% \includegraphics[width=0.4\linewidth]{images/fig_filtering}
% \caption{Signal processing process for extracting peaks from the raw fluorescence data.}
% \label{fig:filtered-signal}
% \end{figure}

@@

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
shown in the plot of Figure \ref{fig:hp_curve}:

To filter out small variations in the signal obtained after applying the function $g$ as well as to eliminate negative values, we use the following hard-threshold filter
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding the Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the importance of spikes that occur when there is a high global activity in the network with respect to  spikes that occur during normal activity. Indeed, we have conjectured that when a large part of the network is firing, the rate at which observations are made is not high enough for being able to detect interactions and, that it would therefore be preferrable to lower their importance by changing appropriatly their magnitude. Additionnally, it is well-known that neurons may also spike because of a global high activity (see \cite{stetter2012model}). In such  context, detecting pairwise neuron interactions from the firing activity is meaningless. 
Accordingly, we have chosen to apply to the signal outputted by $h$ the following function  
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}
which effect is to magnify the importance of spikes that occur in case of low global activity. Note that the particular case where $\sum_{j} x^{t}_j$ equals $0$ is solved by setting $w(x^{t}_i) = 1$.

@@

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
partial correlation coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by \cite{}:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse},
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$.%%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, ie. $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measures between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks \cite{xxx}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performances, with respect to the metric used in the Connectomics
challenge.

Practically, the computation of coefficient $p_{i,j}$ using Equation
\ref{ref} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components of $\Sigma$~\citep{bishop2006pattern} (with
$M$ set to $0.8 p$ in our experiments).

@@

\paragraph{Data and evaluation metrics.}

We report here experiments on the \textit{normal}-1,2,3, and 4
datasets provided by the organizers of the Connectomics
challenge. Each of these datasets is obtained from the simulation
\citep{stetter2012model} of different neural networks of 1000 neurons
and about 15000 edges each (ie., a network density of about
1.5\%). Each neuron is described by a calcium fluorescence time series
of length $T=179500$. All inference methods compared here provide a
ranking of all pairs of neurons according to some association
score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are
summarized using the area under the curve, denoted respectively AUROC
and AUPRC. Only the AUROC score was used to rank the challenge
participants but the precision-recall curve has been shown to be a
more sensible metric for network inference, especially when network
density is small (see e.g. \cite{schrynemackers2013protocols}).
Since neurons are not self-connected in the ground truth networks
(ie., $(i, i) \not \in E, \forall i \in V$), we have manually set the
score of such edges to the minimal possible association score before
computing ROC and PR curves.

\paragraph{Evaluation of the method}

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all four networks
using in each case partial correlation with different filtering
functions. Except for the last two rows that uses PCA, the exact
inverse of the covariance matrix was used in each case. These results
clearly show the importance of the filters. AUROC increases in average from
0.77 to 0.93. PCA does not really affect AUROC scores but it improves
significantly AUPRC scores. Taking the average over various parameter
setting gives an improvement of 10\% in both AUROC and AUPRC. The last
row shows the final performance of the method specifically tuned for
the challenge (see the Appendix for all details). Although this tuning
was decisive to obtain the best performance at the challenge, it does
not significantly improve neither AUROC nor AUPRC. Note that tuning
was mainly done on the \textit{normal-1} and \textit{normal-4}
datasets, which explains why these two networks are slightly better
retrieved than the others.

\begin{table}[t]
\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{@{}l *{8}{c}@{}}
\hline
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\hline
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
No  filtering       & 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
$ h \circ g \circ f_1$                 & 0.925 & 0.925 & 0.924 & 0.923 & 0.322 & 0.324 & 0.323 & 0.312\\
$ w \circ h \circ g \circ f_1$               & 0.931 & 0.930 & 0.928 & 0.927 & 0.333 & 0.334 & 0.327 & 0.313\\
+ PCA         & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
Full method &\\
\hline
%Averaging + Directivity & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389\\
Correlation &\\
GTE & \\
GENIE3 &\\
\end{tabular}
\caption{TODO \textcolor{red}{version simple!!}}
\label{tab:tab3}
\end{table}

@@

In the bottom part of Table \ref{tab:comparison}, the proposed
approach is compared with three other methods: standard (Pearson)
correlation (PC), generalized transfer entropy (GTE), and GENIE3. ROC
and PR curves on the \textit{normal-2} network are shown for all
methods in Figure \ref{fig:}. Pearson correlation measures the
unconditional linear (in)dependence between variables and it should
thus not be able to filter out indirect interactions between
neurons. GTE \citep{stetter2012model} was proposed as a baseline for
the challenge. This method builds on Transfer Entropy to measure the
association between two neurons. Unlike our approach, it can predict
the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is a
gene regulatory network inference method that was the best performer
of the DREAM5 challenge \cite{dream5paper}. When transposed to neural
networks, this method uses as a confidence score for the edge going
from neuron $i$ to neuron $j$ the importance score of variable $X_i$
in a Random Forests model trying to predict $X_j$ from all variables
in $X\setminus X_j$. To reduce computing times of this method, we had
to reduce each tree in the Random Forests model to a maximum depth of
3. This constraint potentially severly affects the performance of this
method with respect to the use of fully grown trees. We nevertheless
provide these results for comparison purpose. PC and GENIE3 were both
applied on the same pre-filtered time series as partial
correlation. For GENIE3, we built 10000 trees per neuron and we used
default settings for all other parameters (except for the maximal tree
depth). For GTE, we reproduced the exact same setting (conditioning
level and pre-processing) that was used for the challenge.

%\paragraph{ROC versus PR curves.} While only ROC was , we argue here that PR scores are more appropriate in this context.

@@

%% The proposed approach is compared to Pearson correlation statistics
%% (as described in Section~\ref{sec:inference}), generalized transfer
%% entropy (GTE) and the GENIE3 algorithm.
%% \begin{itemize}
%% \item \textbf{GTE} \citep{stetter2012model}, the baseline of the challenge,
%% measures for all pairs of neurons $i$ and $j$ the reduction of uncertainty in
%% predicting future values of $i$ given the observation of the time series of the
%% neuron $j$ conditionally to the observation of past time series of neuron $i$,
%% whenever the average fluorescence is over a conditioning level. Pre-processing
%% and conditioning level were set to reproduce baseline of the challenge.

%% \item \textbf{GENIE3} \citep{huynhthu2010inferring} translates the inference
%% of the network of $p$ neurons into $p$ supervised learning tasks. For each
%% task, the fluorescence of a neuron is predicted from the fluorescence of all
%% other neurons using a tree based ensemble such as random forest
%% \citep{breiman2001random}. The tree ensemble associates a variable importance
%% score \citep{louppe2013understanding} to each neuron used in the predictions,
%% which is used as a degree of associate. The following parameters are used
%% $10000$ trees, $\log_2{p}$ variables are randomly selected to split tree nodes
%% and the maximal tree depth is constrained to 1, 2 or 3. \textcolor{red}{AJ Which
%% depth is represented on the graph?}
%% \end{itemize}

%% \subsection*{Partial correlation improves over state-of-the-art methods}

Partial correlation clearly outperforms all other methods. The improvement is more important in terms of AUPRC bu by about a factor
of 2 in AUPRC reaching $0.347$ and an increased of $0.03$ in
AUROC. Averaging Partial correlation matrix over a wide of parameters
further improves AUPRC to $0.389$ and improve AUROC by $0.009$.  Two
reasons might explain such differences: (i) only the partial
correlation-based methods and GENI3 at lesser extent distinguish
direct links from indirect one, (ii) simulated networks from the
\citep{stetter2012model} generator or neural networks in general could
be reduced to a linear model after pre-processing. Those results
generalizes over the other datasets see Appendix \textcolor{red}{REF}.

\begin{figure}[t]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{ROC (left) and PR (right) curves on \text{normal-2} with different methods. Area under the curves are reported in the legend for each method.}

%% The weighted average of partial correlation matrix, winning approach
         %% of the challenge, yields better than AUROC and AUPRC than all
         %% other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         %% averaged partial correlation method, partial correlation used
         %% filter $f_1$). Area under the curves are given between $()$
         %% for the corresponding methods.
%}
\label{fig:curves}
\end{figure}

%% \textcolor{red}{Only one GENI3, thus it can't be observed.}
%% As observed on Figure \ref{fig:curves}, the GENIE3 performance increases with
%% deeper trees while being still computationally feasible. The reason is twofold.
%% First, the deeper the tree, the larger is the conditioning. \textcolor{red}{AJ:
%% conditioning?} As it was observed in the comparison between correlation and
%% partial correlation, it screens indirect links in the inferred network.
%% \textcolor{red}{AJ: I don't understand the link with GENIE3} Second, the number
%% of small trees need to be much higher than deep trees in order to observe - and
%% evaluate - all pairs.  However, as suggested by \cite{louppe2014understanding},
%% limiting the maximal depth of trees may help to improve the variable importance
%% computing, i.e. inferring the degree of association. \textcolor{red}{AJ: there
%% is no explanation of why it would be better apart for computational reason.}
%% Therefore a trade-off  of the maximal depth may give the best of both worlds.
%% \textcolor{red}{AJ:  Which tradeoff? Which worlds?}

\subsection*{ROC versus PR curves}

%% The best method only infers the \textbf{undirected} network by giving the same
%% degree of association to both edges $(i,j)$ and $(j,i)$. This is quite
%% surprising considering the problem statement. Trying to direct links could be
%% more penalizing than rewarding. For example, symmetrizing the degree of
%% association matrix does not necessarily degrade AUPRC and AUROC.

%% If the inferred network corresponds to the undirected network, it would
%% already yield high AUROC score for instance $\approx 0.996$ on
%% \textit{normal-1}. Infering edge directivity is likely to be counter-
%% productive or yield small gain. By contrast, the same experiment with AUPRC
%% would yield a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
%% correctly directivity would be more rewarding in AUPRC term.

%% % Interestingly, methods that takes into account directivity, namely GTE and
%% % GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
%% % (in particular see GTE and GTE(undirected) in Figure \ref{fig:pr_curve}
%% % \textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
%% % especially pr-curve. AS:And now?} and Table \ref{tab:tab1} in Appendix \textcolor{red}{REF}).
%% % Trying to detect the link direction is more penalizing than rewarding.

%% The area under the ROC curve is not a consistent metric with respect to the
%% problem statement. A clear hierarchy is also difficult to derive from ROC
%% curves as shown on Figure \ref{fig:roc_curve}. We advice to avoid the AUROC for
%% the assessment of network inference method. An alternative would be to assess
%% network inference using the area under the precision-recall curve (AUPRC).
%% Those conclusions are in line with \cite{schrynemackers2013protocols}.

@@

In this paper, we present the best performing method of the Connectomics challenge. Our
approach is simple and theoretically sounded and made of two main stages: the
signal is first preprocessed with several filters and then we derive the degree
of association matrix by computing the partial correlation statistics matrix.
We also evaluate the method in a more detailed way and perform a comparison
with other inference methods. We finally discuss the relevance of the metric
used in the challenge and we point out that it was not necessary to discover
causal relationships to maximize the ROC AUC. Thus, it may be  promising to
focus on that particular topic in future works.

@@

% We allow the bibliography to be added on top of the 6 pages.
% Supplementary material may be added in the form or a URL pointed to from
% the paper.

\bibliography{references}

@@
